---
title: "MAC Addresses"
output: html_notebook
---

Conduct a more thorough data analysis into these two MAC addresses including determining locations by using data corresponding to both MAC addresses.  Which of these two MAC addresses should be used and which should not be used for RTLS? Which MAC address yields the best prediction of location?  Does using data for both MAC addresses simultaneously yield more, or less, accurate prediction of location? (Note: this portion is derived from Exercise Q.9 in Nolan and Lang.)

```{r}
library(dplyr)
library(ggplot2)
library(tidyr)
library(ggthemes)

```


This sounds like we are trying to compare three different modeling methods for location finding:
1.  Keep 00:0f:a3:39:e1:c0 and to eliminate the data corresponding to MAC address 00:0f:a3:39:dd:cd. 
2.  Drop 00:0f:a3:39:e1:c0 and to keep the data corresponding to MAC address 00:0f:a3:39:dd:cd. 
3.  Keep both

Our evaluation criteria needs to speak to accuracy - which I would interpret as a higher confidence in the predictions - so having a smaller confidence interval band.

We know our location
We don't know the locations of access points
Could probably describe the location of access points 
We know relationships of access points, test locations, signal strengths
So for the missing one could we MLR it based on the known cases?

Signal Strength = angle + distance(x, y)

```{r}
list <- c("00:0f:a3:39:e1:c0", "00:0f:a3:39:dd:cd")

offline %>% mutate(signal = as.numeric(signal)) %>% filter(mac %in% list) %>% ggplot(aes(x=signal)) + geom_histogram() + theme_light() + facet_wrap(~mac, ncol=2)

```
```{r}
offline %>% ggplot(aes(x=posX, y=posY)) + geom_point(alpha=0.05) + theme_light() + facet_grid(~mac)+ geom_jitter()

```


#Keep e1:c0
```{r}


```


#Keep dd:cd
```{r}


```

#Keep Both Data
```{r}


```