---
title: "MAC Addresses"
output: html_notebook
---

Conduct a more thorough data analysis into these two MAC addresses including determining locations by using data corresponding to both MAC addresses.  Which of these two MAC addresses should be used and which should not be used for RTLS? Which MAC address yields the best prediction of location?  Does using data for both MAC addresses simultaneously yield more, or less, accurate prediction of location? (Note: this portion is derived from Exercise Q.9 in Nolan and Lang.)

```{r}
library(dplyr)
library(ggplot2)
library(tidyr)
library(ggthemes)

```


This sounds like we are trying to compare three different modeling methods for location finding:
1.  Keep 00:0f:a3:39:e1:c0 and to eliminate the data corresponding to MAC address 00:0f:a3:39:dd:cd. 
2.  Drop 00:0f:a3:39:e1:c0 and to keep the data corresponding to MAC address 00:0f:a3:39:dd:cd. 
3.  Keep both

Our evaluation criteria needs to speak to accuracy - which I would interpret as a higher confidence in the predictions - so having a smaller confidence interval band.

We know our location
We figured out the location of the access points.  Two are at the same location.  One was left out of the analysis.
Could probably describe the location of access points 
We know relationships of access points, test locations, signal strengths


Signal Strength = angle + distance(x, y)

```{r}
list <- c("00:0f:a3:39:e1:c0", "00:0f:a3:39:dd:cd")

offline %>% mutate(signal = as.numeric(signal)) %>% filter(mac %in% list) %>% ggplot(aes(x=signal)) + geom_histogram() + theme_light() + facet_wrap(~mac, ncol=2)

```
```{r}
offline %>% ggplot(aes(x=posX, y=posY)) + geom_point(alpha=0.05) + theme_light() + facet_grid(~mac)+ geom_jitter()

```
#Heatmap of each of the two options
```{r}

parCur = par(mfrow = c(4,2), mar = rep(1, 4)) 

mapply(surfaceSS, mac = "00:0f:a3:39:e1:c0", angle = rep(c(0, 135,180, 270), 2), data = list(data = offlineSummary))

par(parCur)

#Nolan, Deborah; Lang, Duncan Temple. Data Science in R: A Case Studies Approach to Computational Reasoning and Problem Solving (Chapman & Hall/CRC The R Series) (Page 28). CRC Press. Kindle Edition. 

```

```{r}
#This is the one that was dropped
compareMacList <- c("00:0f:a3:39:dd:cd", "00:0f:a3:39:e1:c0")
offlineSummary2 = do.call("rbind", signalSummary)     

parCur = par(mfrow = c(2,2), mar = rep(1, 4)) 

mapply(surfaceSS, mac = compareMacList[rep(c(1,2), each=2)], angle = rep(c(90, 135), 2), data = list(data = offlineSummary2))

par(parCur)

#surfaceSS(mac="00:0f:a3:39:e1:c0", angle = 180, data = offlineSummary2)

#Nolan, Deborah; Lang, Duncan Temple. Data Science in R: A Case Studies Approach to Computational Reasoning and Problem Solving (Chapman & Hall/CRC The R Series) (Page 28). CRC Press. Kindle Edition. 

unique(offlineSummary2$orientation)

```
The access point that was eliminated, shown on the bottom, seems to have some greater uncertainty associated with it. While there's a zone of uncertainty associated with the first, it has a higher degree of certainty and strength of signal close to the suspected base.  

#Keep e1:c0
```{r}


```


#Keep dd:cd
```{r}


```

#Keep Both Data
```{r}


```